---
title: "assignmentPS5"
author: "Maansi Surve"
date: "2023-02-17"
output: html_document
---
 R Markdown

 ## --------------------- 1.) Load and check data (5pt)

 You first task is to do a very simple data check:

 1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating

 2. (2pt) Load data. How many rows/columns do we have?

```{r message=FALSE}
  library(tidyverse)
  library(readr)
  library(ggplot2)
  file <- read_delim("gapminder.csv")
```

```{r message=FALSE}
  file %>%
    dim()
```

 3. (2pt) Print a small sample of data. Does it look OK?
```{r message=FALSE}
  file %>%
    sample_n(5)
```

## --------------------- 2 Descriptive statistics (15pt)

 1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.
```{r message=FALSE}
file %>%
  summarize(uniqueByIso3 = n_distinct(iso3), uniqueByIso2 = n_distinct(iso2), 
            uniqueByName = n_distinct(name))
```

 2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are
 even more iso3 -codes. What is going on? Can you find it out?
```{r message=FALSE}
print("There are NA values or empty values being accounted for, which is messing up the numbers.")
```

        (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
        correspond to more than one name? What are these countries?
```{r message=FALSE}
file %>%
  filter(!is.na(iso2), !is.na(name))  %>%
  group_by(iso2)  %>%
  summarize(num = n_distinct(name)) %>%
  arrange(desc(num))
```

        (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
        more than one iso3-code? What are these countries?
        Hint: two of these entities are CHANISL and NLD CURACAO.
```{r message=FALSE}
file %>%
  filter(!is.na(iso3), !is.na(name)) %>%
  group_by(name)  %>%
  summarize(num = n_distinct(iso3)) %>%
  arrange(desc(num))
```

 3. (2pt) What is the minimum and maximum year in these data?  
```{r message=FALSE}
  file %>%
    filter(!is.na(time)) %>%
    summarize(minYear = min(time), maxYear = max(time))
```
 

## --------------------- 3 CO2 emissions (30pt)
 Next, let’s analyze CO2 emissions.
 1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
 and co2_PC. Which years have most missing data?
```{r message=FALSE}
missingCO2 <- file %>%
  group_by(time) %>%
  filter(is.na(co2_PC), is.na(co2)) %>%
  summarize(total = sum(which(is.na(co2_PC)) + which(is.na(co2)))) %>%
  arrange(rank(desc(total)))

print(missingCO2)
```
 
```{r message=FALSE}
missingCO2 %>%
  top_n(10)
```

 2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
 more countries of your choice. Explain what do you see.
```{r message=FALSE}
overtime <- file %>%
  filter(name == "China" | name == "India" | name == "United States of America" | name == "Aruba" | name == "Angola", !is.na(co2), !is.na(name)) %>%
  select(name, time, co2) %>%
  group_by(time) 
  
ggplot(overtime, aes(x = time, y = co2, col = name)) +
  geom_line() +
  xlab("Year") + 
  ylab("CO2 Emissions") +
  ggtitle("CO2 Emissions Over Time (by Country)")
```
 
 
 3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the
 same countries. What does this figure suggest?

```{r message=FALSE}
by_country <- file %>%
  filter(name == "China" | name == "India" | name == "United States of America" | name == "Aruba" | name == "Angola") %>%
  group_by(name, time) %>%
  summarize(co2_pc = sum(co2_PC, na.rm=TRUE))

ggplot(by_country, aes(x = time, y = co2_pc, col = name)) +
    geom_line() +
    xlab("Year") + 
    ylab("CO2 Emissions per Capita") +
    ggtitle("CO2 Emissions per Capita (by Country)")
```

 4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the
 same as continent). Comment what do you see.
```{r message=FALSE}
file %>%
  filter(time %in% c(1960, 2016), !is.na(region), !is.na(co2_PC)) %>%
  group_by(time, region) %>%
  summarize(avgCO2_PC = mean(co2_PC)) %>%
  ggplot(aes(x = region, y = avgCO2_PC, fill=factor(time))) +
  geom_col(position = "dodge") +
    labs(title = "Average Yearly CO2 Emissions Per Capita Across Continents", 
         x = "Continent", y = "Avg CO2 Emissions per Capita") + 
    scale_fill_discrete(name = "Year")
```
 
 Note: just compute averages over countries and ignore the fact that countries are of different size.
    Hint: Americas 2016 should be 4.80.

 5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016.
    Hint: it should look something along these lines:
```{r message=FALSE}
file %>%
  filter(!is.na(region), !is.na(co2), time == 1960 | time == 2016) %>%
  filter(region != "") %>%
  group_by(region, time) %>%
  mutate(avgCO2_PC = mean(co2_PC)) %>%
  ggplot(aes(x = region, y = avgCO2_PC, fill = factor(time))) +
  geom_bar(stat = 'identity', position = 'dodge')
```

 6. Which countries are the three largest, and three smallest CO2 emitters (in terms of   CO2 per capita) in 2019 for each continent? (Assume region is continent). 
```{r message=FALSE}
file %>%
  filter(!is.na(co2_PC), !is.na(name),!is.na(region), time == "2016") %>%
  group_by(region, name) %>%
  summarize(avg_CO2 = mean(co2_PC)) %>%
  arrange(rank(desc(region))) %>%
  group_by(region) %>%
  select(region, name, avg_CO2) %>%
  top_n(3) 
```

```{r message=FALSE}
file %>%
  filter(!is.na(co2_PC), !is.na(name),!is.na(region), time == "2016") %>%
  group_by(region, name) %>%
  summarize(avg_CO2 = mean(co2_PC)) %>%
  arrange(rank(desc(region))) %>%
  group_by(region) %>%
  select(region, name, avg_CO2) %>%
  top_n(-3) 

print(" I went with 2016 since 2019 data is missing.")
```

## --------------------- 4 GDP per capita (50pt)
 Let’s look at GDP per capita (GDP_PC ).

 1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for
 1960. Make the point size dependent on the country size, and color those according to the
 continent. Feel free to adjust the plot in other ways to make it better.
 Comment what do you see there.
```{r message=FALSE}
variables <- file %>%
  filter(time == 1960, !is.na(GDP_PC), !is.na(lifeExpectancy)) %>%
  group_by(GDP_PC, lifeExpectancy) %>% 
  summarize(x = GDP_PC, y = lifeExpectancy, z = region, a = totalPopulation) 

ggplot(variables, aes(x = GDP_PC, y = lifeExpectancy)) +
   geom_point(aes(color = z, size = a, alpha = 0.5)) +
  labs(title = "GDP Per Capita vs. Life Expectancy by Country in 1960", 
         x = "GDP Per Capita", y = "Life Expectancy by Country") +
         guides(color = guide_legend(title = "Continent"), size = guide_legend(title = "Population Size"), alpha = "none")
```

 2. (4pt) Make a similar plot, but this time use 2019 data only.
```{r message=FALSE}
variables <- file %>%
  filter(time == 2019, !is.na(GDP_PC), !is.na(lifeExpectancy)) %>%
  group_by(GDP_PC, lifeExpectancy) %>% 
  summarize(x = GDP_PC, y = lifeExpectancy, z = region, a = totalPopulation) 

ggplot(variables, aes(x = GDP_PC, y = lifeExpectancy)) +
   geom_point(aes(color = z, size = a, alpha = 0.5)) +
  labs(title = "GDP Per Capita vs. Life Expectancy by Country in 2019", 
         x = "GDP Per Capita", y = "Life Expectancy by Country") +
         guides(color = guide_legend(title = "Continent"), size = guide_legend(title = "Population Size"), alpha = "none")
```
 

 3. (6pt) Compare these two plots and comment what do you see. How has world developed
 through the last 60 years?
 
 4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?
    Note: here as average I mean just average over countries, ignore the fact that countries are of different size.
```{r message=FALSE}
file %>%
  group_by(region) %>%
  filter(time == 1960, !is.na(lifeExpectancy), !is.na(region)) %>%
  summarize(avg_lifeExpectancy = mean(lifeExpectancy))
```

```{r message=FALSE}
file %>%
  group_by(region) %>%
  filter(time == 2019, !is.na(lifeExpectancy), !is.na(region)) %>%
  summarize(avg_lifeExpectancy = mean(lifeExpectancy))
```

 5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
 in the order of growth. Explain what do you see.
    Hint: these data (data in long form) is not the simplest to compute growth. But you may
    want to check out the lag() function. And do not forget to group data by continent when
    using lag(), otherwise your results will be messed up! See https://faculty.washington.
     edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.
```{r message=FALSE}
file %>%
  group_by(region) %>%
  filter(!is.na(lifeExpectancy), !is.na(region), region != "") %>%  
  summarize(avgLE = mean(lifeExpectancy)) %>%
   mutate(prev = lag(avgLE),  
          growth = avgLE - prev)
``` 
 
 6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
 histograms on the same graph, see how well you can do it!
```{r message=FALSE}
hist <- file %>%
  filter(time == 1960 | time == 2019, !is.na(GDP_PC))

ggplot(hist, aes(x=GDP_PC, fill = as.factor(time)))+
  geom_histogram(position='dodge') +
  guides(fill = guide_legend(title = "Year")) +
  ggtitle("GDP Per Capita for 1960 & 2019") +
  labs(x = "GDP Per Capita", y = "Year")
```
 

 7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.)
 Hint: check out the function rank()!
 Hint2: 17 for 1960.
```{r message=FALSE}
rank1 <- file %>%
  select(name, time, lifeExpectancy) %>%
  filter(time == 1960, !is.na(lifeExpectancy), name != "") %>%
  arrange(rank(desc(lifeExpectancy))) %>%
  summarize(US_Rank = which(name == "United States of America"))
```

```{r message=FALSE}
rank2 <- file %>%
  select(name, time, lifeExpectancy) %>%
  filter(time == 2019, !is.na(lifeExpectancy), name != "") %>%
  arrange(rank(desc(lifeExpectancy))) %>%
  summarize(US_Rank = which(name == "United States of America"))
```

 8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a
 bit. But we also have more countries in 2019–what about the relative rank divided by the
 corresponding number of countries that have LE data in the corresponding year?
    Hint: 0.0904 for 1960.
```{r message=FALSE}
file %>%
  filter(time == 1960, !is.na(lifeExpectancy), name != "") %>%
  summarize(numCountries = n_distinct(name), rank = rank1/numCountries)
```

```{r message=FALSE}
file %>%
  filter(time == 2019, !is.na(lifeExpectancy), name != "") %>%
  summarize(numCountries = n_distinct(name), rank = rank1/numCountries)
```
```

 9. Finally tell us how many hours did you spend on this PS.
```{r message=FALSE}
print("I spent about 20 hours on this assignment.")
```
 
